\normalfont{A bird's-eye view on the habitability of exoplanets via statistical learning techniques}

### 2020-10-01 Manuscript/Manuscript.tex
PUB

\begin{table}[]
\caption{A selected features of the exoplanets dataset provided by PHL @ UPR Arecibo \cite{planet_dataset} accompanied by their measure unit }
\begin{tabular}{@{}lll@{}}
\toprule
Variable name                                                                & Abbreviation & Description/Comments                                                                                      \\ \midrule
Planet name                                                                  & P\_NAME      & \begin{tabular}[c]{@{}l@{}}The planet name according to \\ the international convention\end{tabular}      \\ \midrule
Planet Period                                                                & P\_P         & The planet period (days)                                                                                  \\ \midrule
Stellar temperature                                                          & S\_T         & \begin{tabular}[c]{@{}l@{}}The star effective temperature\\ (K)\end{tabular}                              \\ \midrule
Planet distance                                                              & P\_D         & \begin{tabular}[c]{@{}l@{}}Mean distance between the planet \\ and the host star (AU)\end{tabular}        \\ \midrule
Planet Periastron                                                            & P\_P         & \begin{tabular}[c]{@{}l@{}}The minimum distance between \\ the planet and the host star (AU)\end{tabular} \\ \midrule
Planet apastron                                                              & P\_A         & \begin{tabular}[c]{@{}l@{}}The maximum distance between \\ the planet and host star (AU)\end{tabular}     \\ \midrule
\begin{tabular}[c]{@{}l@{}}Planet effective \\ thermal distance\end{tabular} & P\_D\_E      & (AU)                                                                                                      \\ \midrule
\begin{tabular}[c]{@{}l@{}}Planet mean stellar\\ flux\end{tabular}           & P\_F         & (earth units)                                                                                             \\ \midrule
\begin{tabular}[c]{@{}l@{}}Planet equilibrium \\ temperature\end{tabular}    & P\_T\_E      & An albedo of 0.3 is considered (K)                                                                        \\ \midrule
Star radius estimated                                                        & S\_R\_E      & (solar units)                                                                                             \\ \midrule
Star luminosity                                                              & S\_L         & (solar units)                                                                                             \\ \midrule
Planet Habitability                                                          & P\_H         & 1 habitable, 0 else                                                                                       \\ \midrule
\begin{tabular}[c]{@{}l@{}}Estimated planet \\ radius\end{tabular}           & P\_R         & (earth units)                                                                                             \\ \midrule
\begin{tabular}[c]{@{}l@{}}Estimated planet\\ mass\end{tabular}              & P\_M         & (earth units)                                                                                             \\ \midrule
Star Spectral Type                                                           & S\_S\_T      & \begin{tabular}[c]{@{}l@{}}The star spectral type according\\ to the diagram in Fig. \ref{H_R_diagram} \end{tabular}         \\ \bottomrule
\end{tabular}
\label{features_table}
\end{table}

@@

The first approach that was considered is the classification decision tree: as we can see from Fig \ref{DecisionTree} the main discriminant is the stellar luminosity: this feature discriminate,as one expects, the largest part of the habitable planet from the non-habitable. The nodes $4$, $9$ and $19$ discriminates deeply the habitability on the basis of stellar temperature planet temperature and its main distance from the host star. Note that the mean distance (P\_D) has a less importance: this can be ascribed to the fact that the habitability was extended beyond the "earth zone" by considering habitable planets that, with a particular atmosphere as shown in fig \ref{Planets_habitability_Seager}, have the correct surface temperature for liquid water. 
\footnote{In principle one may wonder if a better approach should considered only the planets like Earth habitable or perhaps to distinguish the two cases in order to get the distance effect: the issue here is that these exoplanets are very few, thus to preserve the significance of the habitable planets, if such approach was undertaken, a further reduction of the dataset had to be considered. Since this choice strongly limits the quality of forecasts, such approach was not considered.}. The performance of this algorithm on this dataset, evaluated with a 10-fold cross-validation, is reported in Fig \ref{Decision_tree_performances} and in the table \ref{DecisionTree_perf_tab}. Next the random forest approach was considered. As shown in Fig. \ref{Random_forest_trees} the main parameters for this parameters were previously checked. The results of the random forest algorithm are reported in Fig

@@

\begin{figure}[h]
  \centering
\subfloat[]{\includegraphics[width=0.5\textwidth]{Pic/Random_forest_trees.pdf}}
\subfloat[]{\includegraphics[width=0.5\textwidth]{Pic/mtry_Random_forest.pdf}}
\caption{The performances of the random forest algorithm: as the ensemble grew (panel a) accompanied by the optimization of the number of variables available (mtry). For the calculations presented in this work an ensemble of 500 trees (the default one) was considered while the mytry parameter was set equal to 3.}
\label{Random_forest_trees}

>>>>

\begin{table}[]
\caption{A selected features of the exoplanets dataset provided by PHL @ UPR Arecibo \cite{planet_dataset} accompanied by their measure unit }
\begin{tabular}{@{}lll@{}}
\toprule
Variable name                                                                & Abbreviation & Description/Comments                                                                                      \\ \midrule
Planet name                                                                  & P\_NAME      & \begin{tabular}[c]{@{}l@{}}The planet name according to \\ the international convention\end{tabular}      \\ \midrule
Planet Period                                                                & P\_P         & The planet period (days)                                                                                  \\ \midrule
Stellar temperature                                                          & S\_T         & \begin{tabular}[c]{@{}l@{}}The star effective temperature\\ (K)\end{tabular}                              \\ \midrule
Planet distance                                                              & P\_D         & \begin{tabular}[c]{@{}l@{}}Mean distance between the planet \\ and the host star (AU)\end{tabular}        \\ \midrule
Planet Periastron                                                            & P\_PN         & \begin{tabular}[c]{@{}l@{}}The minimum distance between \\ the planet and the host star (AU)\end{tabular} \\ \midrule
Planet apastron                                                              & P\_A         & \begin{tabular}[c]{@{}l@{}}The maximum distance between \\ the planet and host star (AU)\end{tabular}     \\ \midrule
\begin{tabular}[c]{@{}l@{}}Planet effective \\ thermal distance\end{tabular} & P\_D\_E      & (AU)                                                                                                      \\ \midrule
\begin{tabular}[c]{@{}l@{}}Planet mean stellar\\ flux\end{tabular}           & P\_F         & (earth units)                                                                                             \\ \midrule
\begin{tabular}[c]{@{}l@{}}Planet equilibrium \\ temperature\end{tabular}    & P\_T\_E      & An albedo of 0.3 is considered (K)                                                                        \\ \midrule
Star radius estimated                                                        & S\_R\_E      & (solar units)                                                                                             \\ \midrule
Star luminosity                                                              & S\_L         & (solar units)                                                                                             \\ \midrule
Planet Habitability                                                          & P\_H         & 1 habitable, 0 else                                                                                       \\ \midrule
\begin{tabular}[c]{@{}l@{}}Estimated planet \\ radius\end{tabular}           & P\_R         & (earth units)                                                                                             \\ \midrule
\begin{tabular}[c]{@{}l@{}}Estimated planet\\ mass\end{tabular}              & P\_M         & (earth units)                                                                                             \\ \midrule
Star Spectral Type                                                           & S\_S\_T      & \begin{tabular}[c]{@{}l@{}}The star spectral type according\\ to the diagram in Fig. \ref{H_R_diagram} \end{tabular}         \\ \bottomrule
\end{tabular}
\label{features_table}
\end{table}

@@

The first approach that was considered is the classification decision tree. First the complexity parameter (CP) was checked (a 10-fold cross-validation was undertaken): as can be seen from Fig \ref{Decision_tree_performances} the default parameter $0.01$ can be used. Then the decision tree algorithm was applied to the dataset: as it can be seen from Fig \ref{DecisionTree} the main discriminant is the stellar luminosity: this feature discriminate, as one expects, the largest part of the habitable planet from the non-habitable. The nodes $2$, $4$, $9$, $16$, $33$ and $67$ discriminates deeply the habitability on the basis of stellar temperature ($S_T$), the planet distance from the host star ($P\_D$ mean distance and $P\_PN$ periastron) and the planet temperature equilibrium temperature ($P\_T\_E$). Note that the mean distance and the periastron has a less importance with respect to the stellar features: this can be ascribed to the fact that the habitability was extended beyond the "earth zone" by considering habitable planets that, with a particular atmosphere as shown in fig \ref{Planets_habitability_Seager}, have the correct surface temperature for liquid water
\footnote{In principle one may wonder if a better approach should considered only the planets like Earth habitable or perhaps to distinguish the two cases in order to get the distance effect: the issue here is that these exoplanets are very few, thus to preserve the significance of the habitable planets, if such approach was undertaken, a further reduction of the dataset had to be considered. Since this choice strongly limits the quality of forecasts, such approach was not considered.}. The performance of this algorithm on this dataset, evaluated with a 10-fold cross-validation, is reported in Fig \ref{Decision_tree_performances} and in the table \ref{DecisionTree_perf_tab}. Next the random forest approach was undertaken. As shown in Fig. \ref{Random_forest_tree_perfs} the main parameters for this algorithm were previously checked. Its results are reported in Fig

@@

\begin{figure}[h]
  \centering
\subfloat[]{\includegraphics[width=0.5\textwidth]{Pic/Random_forest_trees.pdf}}
\subfloat[]{\includegraphics[width=0.5\textwidth]{Pic/mtry_Random_forest.pdf}}
\caption{The performances of the random forest algorithm: as the ensemble grew (panel a) accompanied by the optimization of the number of variables available (mtry). For the calculations presented in this work an ensemble of 500 trees (the default one) was considered while the mytry parameter was set equal to 3.}
\label{Random_forest_tree_perfs}
